---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "man/figures/README-"
)
```

# cdCAT

<!-- badges: start -->
[![R-CMD-check](https://github.com/thiagofmiranda/cdCAT/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/thiagofmiranda/cdCAT/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

> Computerized Adaptive Testing with Cognitive Diagnostic Models

`cdCAT` is an R package that provides a session-based engine for
**Cognitive Diagnostic Computerized Adaptive Testing (CD-CAT)**.
Unlike simulation-focused packages, `cdCAT` is designed for
**real-time, item-by-item adaptive applications** -- making it suitable
for integration with APIs, Shiny apps, and online assessment platforms.

---

## Theoretical Framework

`cdCAT` implements adaptive testing under **Cognitive Diagnosis Models (CDMs)**.

The adaptive cycle follows the standard CD-CAT framework:

1. Posterior estimation of latent attribute profiles
2. Item selection based on information criteria
3. Response update
4. Stopping rule evaluation

Implemented criteria are grounded in the CD-CAT literature:

- Xu, Chang & Douglas (2003) -- KL
- Cheng (2009) -- PWKL / MPWKL
- Shannon (1948) -- Entropy
- Tatsuoka (2002) -- Attribute-level classification
- Hsu, Wang & Chen (2013) -- Dual posterior threshold
- Kingsbury & Zara (1991) -- Content balancing
- Sympson & Hetter (1985) -- Exposure control
- van der Linden (2005) -- Shadow CAT

This ensures theoretical consistency with established CD-CAT methodology.

---

## Features

- **Models**: DINA, DINO, and GDINA
- **Item selection criteria**: KL, PWKL, MPWKL, SHE, sequential, random
- **Estimation methods**: MAP, MLE, EAP
- **Custom prior**: informative prior over skill profiles via `cdcat_prior()`
- **Fixed or adaptive start item**: control the first item via `start_item`
- **Content balancing**: blueprint-driven domain proportions (Kingsbury & Zara, 1991)
- **Exposure control**: Sympson-Hetter and Randomesque methods
- **Shadow CAT**: user-defined constraint functions for arbitrary test assembly
- **Stopping rules**: maximum items, posterior threshold classification
- **Built-in simulated datasets**: `cdcat_sim` — 16 datasets across K, model, and item quality
- **Session-based architecture**: stateful R6 object manages the full CAT lifecycle
- **Lightweight**: depends only on `R6`

---

## Architecture

`cdCAT` is built around a stateful `CdcatSession` R6 object that:

- Stores the full posterior distribution
- Tracks administered items
- Updates responses sequentially
- Evaluates stopping criteria at each step

The item selection pipeline at each step is:

```
Content balancing filter  ->  Criterion scores  ->  Exposure control  ->  Selected item
                                    |
                          (Shadow CAT replaces the full pipeline
                           when constr_fun is supplied)
```

This design makes the package suitable for:

- REST APIs
- Shiny applications
- Online assessment engines
- Real-time adaptive testing systems

---

## Installation

```{r install, eval=FALSE}
# Development version from GitHub
remotes::install_github("thiagofmiranda/cdCAT")
```

---

## Quick Start

```{r quickstart}
library(cdCAT)

# 1. Define the item bank
Q <- matrix(c(
  1, 0,
  0, 1,
  1, 0,
  0, 1,
  1, 1
), nrow = 5, ncol = 2, byrow = TRUE)

items <- cdcat_items(
  q_matrix = Q,
  model    = "DINA",
  slip     = c(0.10, 0.10, 0.15, 0.10, 0.10),
  guess    = c(0.20, 0.20, 0.15, 0.20, 0.15)
)

# 2. Start a session
session <- CdcatSession$new(
  items     = items,
  criterion = "PWKL",
  max_items = 5L,
  threshold = 0.8
)

# 3. Administer items
simulated_responses <- c(1, 1, 0, 1, 0)

repeat {
  item <- session$next_item()
  if (item == 0) break
  session$update(item, simulated_responses[item])
}

# 4. Results
res <- session$result()
cat("Estimated profile:", res$alpha_hat, "\n")
cat("Items used       :", res$n_items, "\n")
cat("Stop reason      :", res$stop_reason, "\n")
```

---

## Item Selection Criteria

All adaptive criteria operate on the conditional response distribution
P(X_j = 1 | alpha) and use one of the following strategies:

| Criterion  | Full Name                                                   |
|------------|-------------------------------------------------------------|
| `"KL"`     | Kullback-Leibler Information (KL) Method                    |
| `"PWKL"`   | Posterior-Weighted Kullback-Leibler (PWKL) Method           |
| `"MPWKL"`  | Modified Posterior-Weighted Kullback-Leibler (MPWKL) Method |
| `"SHE"`    | Shannon Entropy (SHE) Method                                |
| `"SEQ"`    | Sequential (non-adaptive)                                   |
| `"RANDOM"` | Random selection                                            |

---

## Supported Models

| Model     | Description                         |
|-----------|-------------------------------------|
| `"DINA"`  | Deterministic Input, Noisy And gate |
| `"DINO"`  | Deterministic Input, Noisy Or gate  |
| `"GDINA"` | Generalized DINA                    |

---

## Advanced Features

### Custom Prior

Specify an informative prior over skill profiles using `cdcat_prior()`.
Pattern names follow `expand.grid` order (first attribute varies fastest).

```{r prior, eval=FALSE}
prior <- cdcat_prior(
  "00" = 0.4,
  "10" = 0.3,
  "01" = 0.2,
  "11" = 0.1
)

session <- CdcatSession$new(items, prior = prior)
```

### Content Balancing

Enforce a test blueprint by restricting each selection step to the most
under-represented content domain (Kingsbury & Zara, 1991).

```{r content, eval=FALSE}
content      <- c("Algebra", "Algebra", "Geometry", "Geometry", "Mixed")
content_prop <- c(Algebra = 0.4, Geometry = 0.4, Mixed = 0.2)

session <- CdcatSession$new(
  items        = items,
  content      = content,
  content_prop = content_prop
)
```

### Exposure Control

Limit overuse of specific items with two detection-free methods:

```{r exposure, eval=FALSE}
# Sympson-Hetter: per-item acceptance probabilities in [0, 1]
exposure_sh <- c(0.9, 0.9, 0.9, 0.5, 0.9)   # item 4 accepted only 50% of the time

# Randomesque: top-n pool size per test position (all values >= 1)
exposure_rq <- c(3L, 2L, 2L, 1L, 1L)         # position 1 draws from top-3

session <- CdcatSession$new(items, exposure = exposure_sh)
```

### Fixed or Adaptive Start Item

Control which item is administered first without touching the main
adaptive pipeline. Accepts a fixed index, `"random"`, `"seq"`, or any
adaptive criterion name.

```{r start_item, eval=FALSE}
# Always start with item 3
session <- CdcatSession$new(items, start_item = 3L)

# Draw the first item randomly
session <- CdcatSession$new(items, start_item = "random")

# Use KL for the first item, then switch to PWKL
session <- CdcatSession$new(items, criterion = "PWKL", start_item = "KL")
```

### Shadow CAT

Enforce arbitrary test assembly constraints via a user-supplied function.
`cdCAT` is solver-agnostic: use any optimisation library inside `constr_fun`.

```{r shadow, eval=FALSE}
# constr_fun receives (scores, items, administered) and returns one item index
greedy_shadow <- function(scores, items, administered) {
  scores[administered] <- -Inf
  which.max(scores)
}

session <- CdcatSession$new(items, constr_fun = greedy_shadow)
```

See `vignette("advanced-item-selection")` for a full LP-based example using `lpSolve`.

### Built-in Simulated Data

`cdcat_sim` provides 16 ready-to-use datasets covering K ∈ {3, 5}
attributes, DINA and GDINA models, and four levels of item quality
(HD-LV, HD-HV, LD-LV, LD-HV). Each dataset contains the Q-matrix,
true attribute profiles, item parameters, and simulated responses.

```{r simdata, eval=FALSE}
# List all available datasets
names(cdcat_sim)

# Use the K=3 GDINA high-discrimination dataset
d     <- cdcat_sim[["k3_GDINA_HDLV"]]
items <- cdcat_items(d$Q, "GDINA", gdina_params = d$parameters)
```

See `vignette("simulation-study")` for a worked simulation experiment
comparing all item selection criteria and stopping rules.

---

## How `cdCAT` Differs from Existing Packages

Unlike simulation-oriented packages such as `mirtCAT` or `GDINA`,
`cdCAT` focuses on:

- Real-time session management
- Modular adaptive components (prior, content, exposure, shadow independently composable)
- Lightweight dependency structure (only `R6`)
- Explicit posterior tracking at every step
- Solver-agnostic shadow CAT (no `lpSolve` dependency in the package core)

---

## Roadmap

Planned extensions include:

- Polytomous extensions
- pkgdown documentation site

---

## Authors

- **Thiago Miranda** -- thiagofm.pa@gmail.com
- **Bruno Silvestre** -- bruno.ap.silvestre@gmail.com

---

## License

MIT
