---
title: "Getting Started with cdCAT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with cdCAT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cdCAT)
```

## Overview

`cdCAT` provides a session-based engine for Cognitive Diagnostic 
Computerized Adaptive Testing (CD-CAT). It supports DINA, DINO, and 
GDINA models with multiple item selection criteria and attribute 
estimation methods.

The typical workflow has four steps:

1. Define the item bank with `cdcat_items()`
2. Start a session with `CdcatSession$new()`
3. Loop: select next item → collect response → update session
4. Extract results with `session$result()`

---

## 1. Defining the Item Bank

The item bank requires a Q-matrix and item parameters.
The Q-matrix is a binary matrix where rows are items and columns are 
attributes. An entry of 1 means the item requires that attribute.

```{r items}
# Q-matrix: 5 items, 2 attributes
Q <- matrix(c(
  1, 0,
  0, 1,
  1, 0,
  0, 1,
  1, 1
), nrow = 5, ncol = 2, byrow = TRUE)

# DINA model parameters
items <- cdcat_items(
  q_matrix = Q,
  model    = "DINA",
  slip     = c(0.10, 0.10, 0.15, 0.10, 0.10),
  guess    = c(0.20, 0.20, 0.15, 0.20, 0.15)
)

print(items)
```

---

## 2. Running a CD-CAT Session

```{r session}
# Start session
session <- CdcatSession$new(
  items     = items,
  method    = "MAP",
  criterion = "PWKL",
  min_items = 2L,
  max_items = 5L,
  threshold = 0.8
)

# Simulate responses (1 = correct, 0 = incorrect)
simulated_responses <- c(1, 1, 0, 1, 0)

repeat {
  item <- session$next_item()
  if (item == 0) break
  session$update(item, simulated_responses[item])
}
```

---

## 3. Extracting Results

```{r results}
res <- session$result()

cat("Estimated profile :", res$alpha_hat, "\n")
cat("Items administered:", res$administered, "\n")
cat("Responses         :", res$responses, "\n")
cat("N items           :", res$n_items, "\n")
cat("Stop reason       :", res$stop_reason, "\n")
cat("Posterior         :", round(res$posterior, 3), "\n")
```

---

## 4. Item Selection Criteria

All adaptive criteria below are designed for **single attribute profile
estimation** — they select the item that best discriminates the full
latent attribute profile α.

| Criterion  | Full Name                                                   |
|------------|-------------------------------------------------------------|
| `"KL"`     | Kullback-Leibler Information (KL) Method                    |
| `"PWKL"`   | Posterior-Weighted Kullback-Leibler (PWKL) Method           |
| `"MPWKL"`  | Modified Posterior-Weighted Kullback-Leibler (MPWKL) Method |
| `"SHE"`    | Shannon Entropy (SHE) Method                                |
| `"SEQ"`    | Sequential (non-adaptive)                                   |
| `"RANDOM"` | Random selection                                            |

---

## 5. Supported Models

```{r models, eval=FALSE}
# DINO model
items_dino <- cdcat_items(Q, "DINO", slip = slip, guess = guess)

# GDINA model
gdina_params <- list(
  list("0" = 0.1, "1" = 0.9),
  list("0" = 0.1, "1" = 0.9),
  list("00" = 0.1, "10" = 0.5, "01" = 0.5, "11" = 0.9)
)
items_gdina <- cdcat_items(Q[1:3, ], "GDINA", gdina_params = gdina_params)
```

---

## 6. Estimation Methods

Three attribute estimation methods are available via the `method` argument:

- `"MAP"` — Maximum A Posteriori (default)
- `"MLE"` — Maximum Likelihood Estimation
- `"EAP"` — Expected A Posteriori
